\section{Examples}

The surface system is extremely expressive.
Several example \slang{} constructions can be found in \Fref{surface-examples}.
Turnstile notion is abused slightly so that examples can be indexed by other expressions that obey type rules.
For instance, we can say $refl_{2_{c}:\mathbb{N}_{c}}\ :\ 2_{c}\doteq_{\mathbb{N}_{c}}2_{c}$ since $\mathbb{N}_{c}:\star$ and $2_{c}:\mathbb{N}_{c}$.

\begin{sidewaysfigure}
\begin{tabular}{lllll}
  & $\vdash\perp_{c}$ & $\coloneqq\left(X:\star\right)\rightarrow X$ & $:\star$ & \makecell[l]{Void, ``empty'' type,\\ logical false}\tabularnewline
  & $\vdash Unit_{c}$ & $\coloneqq\left(X:\star\right)\rightarrow X\rightarrow X$ & $:\star$ & Unit, logical true\tabularnewline
  & $\vdash tt_{c}$ & $\coloneqq\lambda-\,x\Rightarrow x$ & $:Unit_{c}$ & \makecell[l]{trivial proposition,\\ polymorphic identity}\tabularnewline
  & $\vdash\mathbb{B}_{c}$ & $\coloneqq\left(X:\star\right)\rightarrow X\rightarrow X\rightarrow X$ & $:\star$ & booleans\tabularnewline
  & $\vdash true_{c}$ & $\coloneqq\lambda-\,then\,-\Rightarrow then$ & $:\mathbb{B}_{c}$ & boolean true\tabularnewline
  & $\vdash false_{c}$ & $\coloneqq\lambda-\,-\,else\Rightarrow else$ & $:\mathbb{B}_{c}$ & boolean false\tabularnewline
$x:\mathbb{B}_{c}$ & $\vdash!_{c}x$ & $\coloneqq x\,\mathbb{B}_{c}\,false_{c}\,true_{c}$ & $:\mathbb{B}_{c}$ & boolean not\tabularnewline
$x:\mathbb{B}_{c},y:\mathbb{B}_{c}$ & $\vdash x\,\&_{c}\,y$ & $\coloneqq x\,\mathbb{B}_{c}\,y\,false_{c}$ & $:\mathbb{B}_{c}$ & boolean and\tabularnewline
  & $\vdash\mathbb{N}_{c}$ & $\coloneqq\left(X:\star\right)\rightarrow(X\rightarrow X)\rightarrow X\rightarrow X$ & $:\star$ & natural numbers\tabularnewline
  & $\vdash0_{c}$ & $\coloneqq\lambda-\,-\,z\Rightarrow z$ & $:\mathbb{N}_{c}$ & \tabularnewline
  & $\vdash1_{c}$ & $\coloneqq\lambda-\,s\,z\Rightarrow s\,z$ & $:\mathbb{N}_{c}$ & \tabularnewline
  & $\vdash2_{c}$ & $\coloneqq\lambda-\,s\,z\Rightarrow s\left(s\,z\right)$ & $:\mathbb{N}_{c}$ & \tabularnewline
  & $\vdash n_{c}$ & $\coloneqq\lambda-\,s\,z\Rightarrow s^{n}\,z$ & $:\mathbb{N}_{c}$ & \tabularnewline
$x:\mathbb{N}_{c},y:\mathbb{N}_{c}$ & $\vdash x+_{c}y$ & $\coloneqq\lambda X\,s\,z\Rightarrow x\,X\,s\,\left(y\,X\,s\,z\right)$ & $:\mathbb{N}_{c}$ & \tabularnewline
$X:\star,Y:\star$ & $\vdash X\times_{c}Y$ & $\coloneqq\left(Z:\star\right)\rightarrow(X\rightarrow Y\rightarrow Z)\rightarrow Z$ & $:\star$ & pair, logical and\tabularnewline
$X:\star,Y:\star$ & $\vdash Either_{c}\,X\,Y$ & $\coloneqq\left(Z:\star\right)\rightarrow(X\rightarrow Z)\rightarrow(Y\rightarrow Z)\rightarrow Z$ & $:\star$ & either, logical or\tabularnewline
$X:\star$ & $\vdash\lnot_{c}X$ & $\coloneqq X\rightarrow\perp_{c}$ & $:\star$ & logical negation\tabularnewline
$x:\mathbb{N}_{c}$ & $\vdash Even_{c}\,x$ & $\coloneqq x\,\star\,\left(\lambda x\Rightarrow\lnot_{c}x\right)\,Unit_{c}$ & $:\star$ & $x$ is an even number\tabularnewline
$X:\star,Y:X\rightarrow\star$ & $\vdash\exists_{c}x:X\Rightarrow Y\,x$ & $\coloneqq\left(C:\star\right)\rightarrow\left((x:X)\rightarrow Y\,x\rightarrow C\right)\rightarrow C$ & $:\star$ & \makecell[l]{dependent pair,\\ logical exists}\tabularnewline
$X:\star,x_{1}:X,x_{2}:X$ & $\vdash x_{1}\doteq_{X}x_{2}$ & $\coloneqq\left(C:\left(X\rightarrow\star\right)\right)\rightarrow C\,x_{1}\rightarrow C\,x_{2}$ & $:\star$ & Leibniz equality\tabularnewline
$X:\star,x:X$ & $\vdash refl_{x:X}$ & $\coloneqq\lambda-\,cx\Rightarrow cx$ & $:x\doteq_{X}x$ & reflexivity\tabularnewline
% $X:\star,x_{1}:X,x_{2}:X$ & $\vdash sym_{x_{1},x_{2}:X}$ & $\coloneqq\lambda p\,C\Rightarrow p\left(\lambda x\Rightarrow C\,x\rightarrow C\,x_{1}\right)\,\left(\lambda x\Rightarrow x\right)$ & $:x_{1}\doteq_{X}x_{2}\rightarrow x_{2}\doteq_{X}x_{1}$ & symmetry\tabularnewline
\end{tabular}

\todo[inline]{use tagged union notation}
\todo[inline]{break out eq for a better table}
\todo[inline]{suc, pred as an example of an unpleasant encoding, also citation}
\todo[inline]{trans, cong}
\todo[inline]{list, vec, singleton}

\caption{Example \SLang{} Expressions}
\label{fig:surface-examples}
\end{sidewaysfigure}

\subsection{Church Encodings}

Data types are expressible using Church encodings (in the style of System F).
Church encodings embed the elimination principle of a data type into higher order functions.
For instance, boolean data is eliminated against true and false, two tags with no additional data.
This can also be recognized as the if-then-else construct that is built into most programming languages.
So $\mathbb{B}_{c}$ encodes the possibility of choice between two elements, $true_{c}$ picks the $then$ branch, and $false_{c}$ picks the $else$ branch.

Natural numbers\footnote{
  Called \textbf{church numerals} in this scheme.
} are encodable with two tags, zero and successor.
In this encoding, the successor tag also contains the result of processing the preceding number.
So $\mathbb{N}_{c}$ encodes those two choices, $(X\rightarrow X)$ handles the recursive result of the prior number in the successor case, and the $X$ argument specifies how to handle the base case of $0$.
This can be viewed as a looping construct with temporary storage that loops exactly as many times as the number it represents.

Parameterized data types such as pairs and the $Either_c$ type can also be encoded in this scheme.
A pair type can be used in any way the two terms it contains can, so a pair is defined as the curried input to a function.
The $Either_c$ type is handled if both possibilities are handled, so it is defined as a higher order function that will return an output if both possibilities are handled for input.

\todo{church encoding citation?}

% not necessarily convenient
Church encodings provide a theoretically lightweight way of working with data in a minimal lambda calculus.
However, they are inconvenient.
For instance, the predecessor function on natural numbers is not as simple as it would seem.\todo{note the pulling teeth anticdote?}
To make the system easier for programmers, data types will be added directly in \ch{4}.

\subsection{Proposition Encodings}

\todo{make clear the destinction between bool and LOGIC?}

In general we associate the truth value of a proposition with the inhabitation of a type by a meaningful value.
This meaningful term corresponds to a proof.
So, $\perp_{c}$, the ``empty'' type, can be interpreted as a false proposition, while $Unit_{c}$ can be interpreted as a trivially true proposition, since it has only one good inhabitant\footnote{
  Remember to keep the the different notions of ``truth'' separate:

  $\mathbb{B}$ is a collection of 2 tags $true$ and $false$.
  The names are abitrary, nothing but convention informs their meaning.
  Just as if-then-else constructs could be reordered into if-else-then without changing anything essential.

  In type theory the $\perp$ proposition contains no proofs.
  If you ever get one, something has gone wrong and you have an excuse to do anything.
  Meanwhile the $Unit$ proposition contains just a single trivial proof.
  Nothing interesting can be done with it, just as nothing interesting can be done with the identity function.
  
  Finally, these notions of ``truth'' are distinct from the meta theoretic properties that will be presented later.
}.

Several of the church encoded data types we have seen can also be interpreted as logical predicates.
For instance, the tuple type can be interpreted as logical $and$ since $X\times_{c}Y$ can only be meaningfully inhabited when both $X$ and $Y$ are inhabited.
The $Either$ type can be interpreted as logical $or$ since $Either_{c}\,X\,Y$ can be inhabited when either $X$ or $Y$ is inhabited.

With dependent types, more interesting logical predicates can be encoded.
For instance, we can characterize when a number is even with $Even_{c}$.
We can show that $2$ is even by showing that $Even_{c}\,2_{c}$ is inhabited with the term $\lambda s\Rightarrow s\,tt_{c}$.
Since the definition of $Even_{c}\,2_{c}$ expands to $(Unit_{c} \rightarrow \perp_{c}) \rightarrow \perp_{c}$, given a function $s : (Unit_{c} \rightarrow \perp_{c})$ we only need to give it a member of $Unit_{c}$ to satisfy the type constraint.

Other predicates are encodable in this style (See \cite{Martin-Lof-1971,cardelli1986polymorphic,10.1016/0890-5401(88)90005-3} for more examples).
For instance, we can encode the existential quantifier as $\exists_{c}$ as shown in \Fref{surface-examples}.
Then we can show $\exists_{c}x:\mathbb{N}_{c}\Rightarrow\,Even_{c}\,x$ with a suitable inhabitant of that type.
$0$ is clearly an even number, so our inhabitant could be $\lambda-f\Rightarrow f\,0_{c}\,tt_{c}$, since the $Even_{c}\,0_{c}$ expands to $Unit_{c}$ so $tt_{c}\ :\ Even_{c}\,0_{c}$.
% Note that the existential is equivalent to the tuple if $Y$ does not depend on the value of $X$.

One of the most interesting propositions is the proposition of equality.
$\doteq$ is referred to as \textbf{Leibniz equality} since two terms are equal when they behave the same on all predicates\footnote{
  Originally, Leibniz assumed a metaphysical identification of ``substance'', not a mathematical notion of equality\cite[Section 9]{Leibniz1686}.
  Over time the principle evolved into the current notion of ``the identification of indiscernibles'', and is referred to as \textbf{Leibniz's law}.
}. \todo{cite the ency of philosophy}
We can prove $\doteq$ is an equivalence within the system by proving it is reflexive, symmetric, and transitive.
\todo{give example}
% Additionally we can prove congruence.
\todo[inline]{talk more about congruence}
\todo[inline]{footnote on Leibniz equality for alt encoding}

\subsection{Large Eliminations}

\todo[inline]{double check large elimination def.
consistent with the notes % https://github.com/RobertHarper/hott-notes/blob/5339576f55a4b7f5d04734370a5117491c44b1fe/notes_week5.tex#L155 .
would like better explanation}

It is useful for a type to depend specifically on term level data, this is called \textbf{large elimination}.
Large elimination can be simulated with \tit{}.

\begin{tabular}{llll}
  $toLogic$ & $\coloneqq\lambda b\Rightarrow b\,\star\,Unit_{c}\,\perp_{c}$ & $:$ & $\mathbb{B}_{c}\rightarrow\star$\tabularnewline
  $isPos$ & $\coloneqq\lambda n\Rightarrow n\,\star\,(\lambda-\Rightarrow Unit_{c})\,\perp_{c}$ & $:$ & $\mathbb{N}_{c}\rightarrow\star$\tabularnewline
\end{tabular}
  
For instance, $toLogic$ can convert a $\mathbb{B}_{c}$ term into its corresponding logical type, $toLogic\ true_{c}\,\equiv\, Unit_{c}$ while $toLogic\ false_{c}\, \equiv\, \perp_{c}$.
The expression $isPos$ has similar behavior, going to $\perp_{c}$ at $0_{c}$ and $Unit_{c}$ otherwise.

Note that such functions are not possible in the Calculus of Constructions.

\subsection{Inequalities}

Large eliminations can be used to prove inequalities that can be hard or impossible to express in other minimal dependent type theories. % such as the Calculus of Constructions.
For instance, \todo{ramsy flaged something here, was it the page break?}

\begin{tabular}{lcll}
  $\lambda pr\Rightarrow pr\,\left(\lambda x\Rightarrow x\right)\,\perp_{c}$ & : & $\lnot_{c}(\star\doteq_{\star}\perp_{c})$ & \makecell{The type universe is distinct\\ from Logical False}\tabularnewline
  $\lambda pr\Rightarrow pr\,\left(\lambda x\Rightarrow x\right)\,tt_{c}$ & : & $\lnot_{c}(Unit_{c}\doteq_{\star}\perp_{c})$ &  \makecell{Logical True is distinct\\ from Logical False}\tabularnewline
  $\lambda pr\Rightarrow pr\,toLogic\,tt_{c}$ & : & $\lnot (true_{c}\doteq_{\mathbb{B}_{c}}false_{c})$ &  \makecell{Boolean true and false\\ are distinct}\tabularnewline
  $\lambda pr\Rightarrow pr\,isPos\,tt_{c}$ & : & $\lnot (1_{c}\doteq_{\mathbb{N}_{c}}0_{c})$ & 1 and 0 are distinct\tabularnewline
\end{tabular}
  
\todo[inline]{1st not sensible in CC}

\todo[inline]{2 possibly in CC?}

Note that a proof of $\lnot1_{c}\doteq_{\mathbb{N}_{c}}0_{c}$ is not possible in the Calculus of Constructions\cite{10.2307/2274575}\footnote{
  Martin Hofmann excellently motivates the reasoning in the exercises of \cite{hofmann_1997}.
  % For any encoding of Nats [Guevers?]
}.
\todo[inline]{citation is actually for an MLTT, but if good enough for Hoff good enough for me}

\subsection{Recursion}

The syntax of functions contain a variable to perform in unrestricted recursion.
Though not always necessary\footnote{
  For instance, Church numerals have a limited form of recursive behavior built in.
}, recursion can be very helpful for writing programs.
For instance, here is (an inefficient) function that calculates Fibonacci numbers.

$\mathsf{fun}\,f\,x\Rightarrow case_{c}\,x\,0_{c}\left(\lambda px\Rightarrow case_{c}\,px\,1_{c}\left(\lambda-\Rightarrow f\left(x-_{c}1\right)+_{c}f\left(x-_{c}2\right)\right)\right)$

Assuming appropriate definitions for $case_{c}$, and subtraction.

Recursion can also be used to simulate induction. 
We will not see much of recursion until \ch{4}, when data types are introduced and larger examples are easier to express.
\todo[inline]{is predecessor a better example?}
\todo[inline]{GCD, recursive types?}