\chapter{Notes and Future Work}
\label{chapter:Notes}
\thispagestyle{myheadings}
 
The content of this thesis was achieved through much trial and error.
There were several experiments that while interesting and promising, did not cleanly fit into the narrative of the first 5 Chapters.
This Chapter contains an idealized\footnote{
  For instance, automated testing procedure was originally specified and implemented on a different language than the \clang{} described in this thesis.
  Accordingly the notations have changed, to be sensible in the context of the rest of the thesis. % , often improving/simplifing the presentation.
  Further, automated testing procedure has not yet been reimplemented on the current version of the prototype.
} review of what was tried, and hopefully provide hints about how one might productively try again.
 
The goal was always to make a dependently typed language as approachable as python\footnote{We have perhaps succeeded in making a dependently typed language as approachable as Haskell.}.
While I believe a \fullSp{} dependently typed language with runtime definitional equality checking, dependent pattern matching and with no restrictions on recursion are part of that practical language, there are still lingering usability issues.
For those who share the dream of more reliable software through easier to use dependently typed languages, here are some lines of work for your consideration.

\input{8_Notes/1_Sym} 
\input{8_Notes/2_Proof_Search}

\section{Future work}

\todo[inline]{Convenience: implicit function arguments}

\subsection{Effects}

The last and biggest hurdle to bring dependent types into a mainstream programming language is by providing a reasonable way to handle effects.
Though dependent types and effects have been studied I am not aware of any \fullSp{} system that has implemented those theories.
It is not even completely clear how best to add an effect system into Haskell, the closest ``mainstream'' language to the one studied here.\todo{cites!!}

While trying carefully to avoid effects in this thesis, we still have encountered 2 important effects: blame based error and non-termination.

\subsubsection{Errors}

The current system implements blame based runtime errors and static warnings in a unique way.
There is no control flow for errors built into the reduction or \cbv{} relations, and there is no way to handle an error within the program.
Every potential error is linked to a static warning.
There are a few features that would be good to experiment with.

Ideally we could allow users to provide proofs of equality to remove warnings by having them define and annotate an appropriate identity type.
This would allow the language to act more like an Extensional Type Theory.
Programmers could justify these proofs as a way to remove runtime checks and make code (and testing) faster.
Just as with \ac{ETT} many desirable properties such as function extensionality would still not be provable.
We have pushed this to future work since there are already many explored strategies for dealing with equality proofs in an Intentional Type Theory that are suitable for avoiding warnings in the current implementation.

Currently blame based errors aren't handled\footnote{or caught}.
Programmers may want to use the information from a bad cast to build the final output, it might even be possible to capture a well typed term that witnesses the inequality.
For instance,
\todo{fix example}
\begin{lstlisting}
f : Vec String 1 -> String;
f x = case x {
  Cons _ a _ _ => a
}

h : (x : Nat) -> String;
h x =
  handle{
    f (rep String "hi" x)
  } pr : x != 1 => "whoops" ;
\end{lstlisting}

Though additional research would be needed for exactly the form the contradictions should take if they are made available to the handler.

Handling effects in a dependent type system is subtle\footnote{everything about dependent types is subtle} since the handling construct can observe differences that should not otherwise be apparent.
This is most clearly seen in the generalization of Herbelin's paradox\todo{independent cite} presented in \cite{pedrot2020fire}.
The problem is that the value of a \Bool{} term may depend on effects that cause logical unsoundness (or worse).
The paradox can be presented in our system with an additional handling construct,

\begin{lstlisting}
h : (u : Unit) -> Bool;
h u =
  handle{
    case u {
      | tt => true
    }
  } _ => false ;

hIsTrue : (u : Unit) -> Id Bool true (h u);
hIsTrue u =
  case u <u -> Id Bool true (h u)>{
    | tt => refl Bool true
  };

hIsTrue !! : Id Bool true false
\end{lstlisting}

Interestingly this term is not as bad as the paradox would be in other settings.
A warning is given so we would not expect logical soundness.
If evaluated in \whnf{} the term will produce blame witnessing the static warning given.

\todo{doe ther errors kind of work like those Error TT papers?} 

\subsubsection{Non-termination}

Non-termination is allowed, but it would be better to have it work in the same framework as equational warnings, namely warn when non-termination is possible, and try to find slow running code via symbolic execution (runtime errors for non termination are famously not possible).
Then we could say without caveat ``programs without warnings are proofs''.
It might be possible for users to supply their own external proofs of termination\cite{casinghino2014combining}, or termination metrics\todo{Ats, ``Termination casts: A flexible approach to termination with general recursion''}.

\subsubsection{Other effects}

One of the difficulties of an effect system for dependent types is expressing the definitional equalities of the effect modality.
Is $\mathtt{print"hello";print"world"}$ $\equiv$ $\mathtt{print"helloworld"}$ at type $\mathtt{IO\ Unit}$?
By delaying equality checks until runtime these issues can be avoided until the research space is better explored.
Effects risk making computation mathematically inelegant.
In this thesis we avoided this inelegance for an error effect with a blame system.
Something analogous could perhaps be applied to more interesting effect systems.

Both the symbolic execution and search above could be considered in terms of an effect in an effect system.
Proof search could be localized though an effect modality, avoiding the non local examples above.

\section{User studies}
 
The main proposition of this work is that it will make dependent types easier to learn and use.
This should be demonstrated empirically with user studies.
 
\section{Semantics}
 
The semantics explored in this thesis have been operational, this has led to serviceable, but cumbersome, proofs.
Ideally less syntactic semantics of a typed language in this style should be explored.
 
For instance, the entire system is designed with an unformalized notion of observational equivalence\todo{foot also called...} in mind.
While there has been some exploration into untyped observational equivalence for dependent types in \cite{sjoberg2015dependently,jia2010dependent}, it uses untyped contextual equivalence, which is a weak relation\todo{
  Ian A. Mason and Carolyn L. Talcott. Equivalence in functional languages
  with effects. Journal of Functional Programming, 1(3):287--327, 1991}.
A version of typed operational equivalence is considered in \cite{VAKAR2018401} though they consider definitional distinctions observable. % (is does notseem reasonable to consider
There is a difficult circularity when trying to define typed observational equivalence in a dependently typed setting.
A good of dependently typed observational equivalence would be an interesting and helpful direction for further study.
