\section{Cast soundness}

For the fragment without data, we can show cast soundness.
Again we conjecture that the proof could be extended to support data.
 
If $\varGamma\vdash a\sqsupseteq a':A'$, $\varGamma\ \mathbf{Empty}$ then either $a\ \mathbf{Consistent}$, $a\ \mathbf{Blame}_{\ell,o}$, or $a\rightsquigarrow b$.
Again progress can be shown by extending the usual \ac{TAS} progress argument over the new constructs.

The $\mathbf{Consistent}$ judgment (\Fref{cast-data-val}) generalizes being a value, to elements over $\cup$ and $\sim$.
This only matters for functions, since they are treated extentionally (so that under applied data constructors can be treated as functions).
This relation can be restricted further to be more like a conventional value judgment, for instance enforcing the arguments of a data type are consistent, but these restrictions are not needed here.

\begin{figure}
\[
\frac{\ }{\star\ \mathbf{Consistent}}
\]

\[
\frac{\ }{\left(x:A\right)\rightarrow B\ \mathbf{Consistent}}
\]

\[
\frac{\ }{\left(\mathsf{fun}\,f\,x\Rightarrow b\right)\ \mathbf{Consistent_{\mathsf{fun}}}}
\]

\[
\frac{
  a\ \mathbf{Consistent_{\mathsf{fun}}}\quad b\ \mathbf{Consistent_{\mathsf{fun}}}
}{
  a\sim_{\ell,o}^{\left(x:A\right)\rightarrow B}b\ \mathbf{Consistent_{\mathsf{fun}}}}
\]

\[
\frac{
  a\ \mathbf{Consistent_{\mathsf{fun}}}\quad b\ \mathbf{Consistent_{\mathsf{fun}}}
}{
  a\cup^{\left(x:A\right)\rightarrow B}b\ \mathbf{Consistent_{\mathsf{fun}}}
}
\]

\[
\frac{
  a\ \mathbf{Consistent_{\mathsf{fun}}}
}{
  a::\left(x:A\right)\rightarrow B\ \mathbf{Consistent_{\mathsf{fun}}}}
\]

\[
\frac{
  \mathbf{length}\,\overline{a} < \mathbf{length}\,\Delta
}{
  D_{\Delta}\,\overline{a}\ \mathbf{Consistent_{\mathsf{fun}}}
}
\]

\[
\frac{
  \mathbf{length}\,\overline{b} < \mathbf{length}\,\Delta
}{
  d_{\Delta\rightarrow D\overline{a}}\,\overline{b}\ \mathbf{Consistent_{\mathsf{fun}}}
}
\]

\[
\frac{
  \mathbf{length}\,\overline{a} = \mathbf{length}\,\Delta
}{
  D_{\Delta}\,\overline{a}\ \mathbf{Consistent}
}
\]

\[
\frac{
  \mathbf{length}\,\overline{b} = \mathbf{length}\,\Delta
}{
  d_{\Delta\rightarrow D\overline{a}}\,\overline{b}\ \mathbf{Consistent}
}
\]

\caption{Consistent}
\label{fig:cast-data-val}
\end{figure}


The $a\ \mathbf{Blame}_{\ell,o}$ judgment means a witness of error can be extracted from the term $a$ pointing to the original source location $\ell$ with observation $o$.
The most important rule is,

\[
\frac{\mathbf{head}\ a\neq\mathbf{head}\ b\quad}{a\sim_{\ell,o}^{L}b\ \mathbf{Blame}_{\ell,o}}
\]

Blame can be recursively extracted out of every sub expression.
For instance, $\left(\left(1\sim_{\ell,app_{1}}^{\Nat{} }0\right)+2\right)\ \mathbf{Blame}_{\ell,app_{1}}$.
% In the implementation type errors are preferred.

There are two new sources of blame from the \case{} construct.
The \clang{} records every ``unhandled'' branch and if a \scrut{} hits one of those branches the \case{} will be blamed for in-exhaustiveness\footnote{
  This runtime error is conventional in ML style languages, and is even how Agda handles incomplete matches.
}.
If a \scrut{} list primitively contradicts the pattern coverage blame can be extracted from the \scrut{}. 
Since our type system will ensure complete coverage (based only on constructors) if a \scrut{} escapes the complete pattern match, it must be that there was blamable a cast in the \scrut.
% We have elided most of the structural rules that extract blame from terms, paths, and casts.
% We have left the structural rule for explicit blame for emphasis.



% \begin{figure}
% \todo[inline]{REDO THIS}

% \[
% \frac{\textbf{Blame}\:\ensuremath{\ell}\,o\ p}{\left(d'\ \overline{patc}\right)::p\ \mathbf{!Match}\ \ensuremath{\ell}\,o\,\left(d\ \overline{patc}\right)::p}
% \]

% \todo[inline]{This figure is not very helpful?}

% \caption{\CLang{} Blame}
% \label{fig:cast-data-NoMatch}
% \end{figure}