\section{Data}

%What data defs/ constructors
% The langugae defined in this thesus uses data definitions like those found in systems like Agda and Coq. 

A dependent data type is defined by a type constructor indexed\footnote{
 In more developed systems such as Coq and Agda data types may also have parameters.
 These are indices that apply uniformly over the type and every term construtor.
 For instance, the first argument of \Vect{} is often given by a parameter.
% \begin{lstlisting}[basicstyle={\ttfamily\tiny}]
% data Vec (A : *) : Nat -> * {
% | Nil  : Vec A Z
% | Cons : A -> (x : Nat) -> Vec A x -> Vec A (S x)
% };
% \end{lstlisting}
 Parameters help remove clutter, help inference and erasure, and presumably have parametric properties in the sense of \cite{10.1145/99370.99404}.
 However parameters are easy to simulate with indexes, so for simplicity we will only deal with \emph{indexed} dependent data.
 Parameters can simulate indexes with a suitable equality type\cite{sjoberg2012irrelevance}, though this is not always possible\cite[Section 8]{10.1145/3495528}.
} by arguments, and a set of data constructors that tag data and refine those arguments.
Several familiar data types are defined in Figure \ref{fig:data-defs}.
For example, the data type of natural numbers is defined with the type constructor \Nat{} (which has no type arguments), the data constructors $\mathtt{Z}$ which takes no further information and the data constructor \Suc{} which is formed with the prior number.
The data type $\mathtt{Vec}$ has two type arguments corresponding to the type contained in the vector and its length; it has two data constructors that allow building an empty vector, or to add an element to the front of an existing vector.

Data defined in this style is simple to build and reason about, since data can only be created from its constructors.
Unfortunately the details of data elimination are a little more involved.

\begin{figure}
\begin{lstlisting}[basicstyle={\ttfamily\small}]
data Void : * {};

data Unit : * {
| tt : Unit
};

data Bool : * {
| True : Bool
| False : Bool
};
 
data Nat : * {
| Z : Nat
| S : Nat -> Nat
};
three : Nat;
three = S (S (S Z)));
-- Syntactic sugar allows 3 = S (S (S Z)))

data Vec : (A : *) -> Nat -> * {
| Nil  : (A : *) -> Vec A Z
| Cons : (A : *) -> A -> (x : Nat)
        -> Vec A x -> Vec A (S x)
};

someBools : Vec Bool 2;
someBools = Cons Bool True 1 (Cons Bool False 0 (Nil Bool));

data Id : (A : *) -> A -> A -> * {
| refl  : (A : *) -> (a : A) -> Id A a a
};

threeEqThree : Id Nat 3 3;
threeEqThree = refl Nat 3;
\end{lstlisting}
% include if notation is used
% -- Syntactic sugar expands list notation,
% -- for example
% -- [True, False]<Bool> =
% -- Cons Bool True 1 (Cons Bool False 1 (Nil Bool))
 \caption{Definitions of Common Data Types}
\label{fig:data-defs}
\end{figure}
